Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
update type
Fix types
minor

This section sets out to propose a candidate design for a type language that
can be used to describe the arguments and return values of functions in R.
The goal is not to be the final design but rather a starting point for an
iterative process.

Fig.~\ref{types} presents our type language. An early design choice was to
stay close to the R language and only depart in small, hopefully,
non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
variable length argument list. The rest of this section details and
motivates our design.

@@

\paragraph{Scalar vs. Vectors} Our type system distinguishes between
vectors of length 1, and vectors of any dimensions. The six base types,
integers, doubles, complex numbers, logicals (booleans), characters, and
raws (bytes), can be either vectors (e.g., \VEC\int) or scalars (e.g.,
\int).  It is worthwhile to note that a vector can happen to be of length 1
and thus a scalar value is a subtype of the corresponding vector type.
Vectors are homogeneous, in that a vector of doubles contains only doubles.

\paragraph{Missing values} In R, a missing value is written
\NA, each of basic types has its specific representation of \NA. Thus there
is an \int missing observation as well as a double missing, and they are
different.  The type system allows to distinguish between vectors that may
contain missing observations (written \NAVEC\int) and those who are
guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be treated as a
vector that may have missing observations.  We do not allow scalar values to
be missings.  The type \raw does not allow {\NA}s.

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{|} \T_n$.

\paragraph{Nullables} 
The type system has a \null type that is inhabited by a singleton
\code{NULL} value. This value is often used as a sentinel or as a stand in
for arguments that were not provided.  To capture this behavior, we
introduce a nullable type \code{?}\T.  The difference between the \NA and
\code{NULL} values is that \NA can be stored in primitive vectors.

@@

\code{dplyr::group_indices}: \FUN{\CLASS{data.frame}, \VARGS}{\VEC\int}

\code{lava::index}: \FUN{\CLASS{lvm} \; | \; \CLASS{lvmfit}, \VARGS}{\; ? \; \LIST{any}}

\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{dip} \; | \; \dbl}

@@

\code{matrixcalc::matrix.power}: \FUN{\CLASS{matrix}, \dbl}{\CLASS{matrix}}

$$type `data.tree::Traverse` \FUN{\CLASS{`Node`, `R6`}, ~\VEC\chr, \any, \any}\LIST\any;$$

$$type `openssl::decrypt\_envelope` \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{`key`, `rsa`}, \any}~\VEC\raw;$$

$$dbplyr:type `set\_win\_current\_group` \FUN{? ~\VEC\chr}? ~\VEC\chr;$$

$$type `sha256` \FUN{~\VEC\raw, ? ~\VEC\raw}~\VEC\raw;$$

$$type `initparam` \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}~\VEC\dbl;$$

>>>>

This section sets out to propose a candidate design for a type language to
describe the arguments and return values of functions in R.  The goal is not
to get a final design but rather a starting point for an iterative process.
Fig.~\ref{types} presents our type language. An early design choice was to
stay close to the R language and only depart in small, hopefully,
non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
variable length argument list. The rest of this section details and
motivates our design.

@@

\paragraph{Scalar vs. Vectors} We distinguish between
vectors of length 1, and vectors of any dimension. The base types, \int,
\dbl, \clx, \lgl (booleans), \chr, and \raw (bytes), can be either vectors
(e.g., \VEC\int) or scalars (e.g., \int).  A vector can happen to be of
length 1, and thus a scalar value is also a vector.  Vectors are
homogeneous, in that a vector of doubles contains only doubles.

\paragraph{NA values} In R, a value can be \emph{not available}, written
\NA. Each of basic types has its specific \NA. Thus there is an \int \NA as
well as a double \NA, and they are different.  The type system allows to
distinguish between vectors that may contain {\NA}s(written \NAVEC\int) and
those who are guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be
treated as a vector that may {\NA}s.  We do not allow scalar values to be
\NA and the type \raw does not allow {\NA}s.

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{~|~} \T_n$.

\paragraph{Nullables} 
The type system has a \null type inhabited by a singleton \code{NULL}
value. This value is often used as a sentinel or as a stand in for arguments
that were not provided.  To capture this behavior, we introduce a nullable
type \code{?}\T.  The difference between the \NA and \code{NULL} values is
that \NA can be stored in primitive vectors.

@@

\code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}

\code{lava::index}: \FUN{\CLASS{\texttt{lvm}} \; | \; \CLASS{\texttt{lvmfit}}, \VARGS}{\; ? \; \LIST{\any}}

\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl}

@@

\code{matrixcalc::matrix.power}: \FUN{\CLASS{\texttt{matrix}}, \dbl}{\CLASS{\texttt{matrix}}}

\code{data.tree::Traverse}: \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr, \any, \any}{\LIST{\any}}

\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}~\VEC\raw

\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}? ~\VEC\chr

\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}~\VEC\raw

\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}~\VEC\dbl
